---
title: "`swaRm` - a package for processing collective movement data"
author: "Simon Garnier - <a href='https://twitter.com/sjmgarnier'>@sjmgarnier</a>"
date: "`r Sys.Date()`"
output:
  html_document:
    css: custom.css
    fig_caption: yes
    fig_height: 6
    fig_width: 8
    number_sections: yes
    theme: cosmo
    highlight: pygments
    toc: yes
  pdf_document:
    toc: yes
vignette: |
  %\VignetteIndexEntry{Vignette Title} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

---

# Introduction

The goal of this vignette is to demonstrate the functionalities of the `swaRm` 
package for processing collective movement data. 

[Back to top]("#")

---

# Installation

`swaRm` can be installed from the [SwarmLab](http://www.theswarmlab.com)'s 
[`drat` repository](https://github.com/swarm-lab/drat) by running the following
lines in your R terminal: 

```{r, eval=FALSE}
install.packages("drat")  # This line is required if drat is not present 
                          # in your R installation.

drat::addRepo("swarm-lab")
install.packages("swaRm")
```

If you want to receive package updates automatically, you can add the `drat` 
repository to your R installation permanently by adding the line 
`drat::addRepo("swarm-lab")` to your `.Rprofile` file.

[Back to top]("#")

---

# Preparing a trajectory data table

The `swaRm` package ships with a number of example data files that we will use 
for demonstrating its functioning. 

First, we will load in memory the first of these example data files. 

```{r}
filePath <- system.file("extdata/01.csv", package = "swaRm")
dat01 <- read.csv(filePath)
head(dat01)
```

This file contains GPS data, with a date, a timestamp, and longitude-latitude 
coordinates. Note that the coordinates can also be more classical X-Y cartesian
coordinates. 

The first step of the analysis process is to transform this imported data into 
a standard trajectory data table that will be usable by all the function in the 
`swaRm` package. 

```{r}
library(swaRm)
traj01 <- makeTraj(x = dat01$lon, y = dat01$lat, id = "01", 
                   date = dat01$date, time = dat01$time, 
                   date.format = "ymd", time.format = "hms",
                   geo = TRUE)  # Set 'geo' to FALSE if working with cartesian
                                # (i.e. non geographic) coordinates
traj01
```

[Back to top]("#")

---

# Dealing with missing observations

Missing observations are frequent in tracking data, especially when using GPS 
that can lose track of satellites frequently. `swaRm` provides convenience 
functions to deal with missing observations by first detecting them, and then
replacing them with interpolated observations (either linearly or using splines).

For instance in the imported data, we are missing a few seconds at the beginning
of the file. We can complete the trajectory as follows. 

```{r}
traj01 <- completeTraj(traj01, geo = TRUE)
traj01

traj01 <- fixTraj(traj01, geo = TRUE) # Use 'spline = TRUE' to interpolate the 
                                      # missing observations with a spline function
                                      # instead of a linear one. 
traj01
```

[Back to top]("#")

---

# Computing the basic parameters of a trajectory

## Linear distances between successive steps

```{r}
traj01[, linDist := linDist(x = lon, y = lat, geo = TRUE)]
traj01[, cumDist := cumsum(linDist)]
traj01
```

## Linear speeds between successive steps

```{r}
traj01[, linSpeed := linSpeed(x = lon, y = lat, t = time, geo = TRUE)]
traj01
```

## Linear accelerations between successive steps

```{r}
traj01[, linAcc := linAcc(x = lon, y = lat, t = time, geo = TRUE)]
traj01
```

## Headings between successive steps

```{r}
traj01[, heading := heading(x = lon, y = lat, geo = TRUE)]
traj01
```

## Angular speeds between successive steps

```{r}
traj01[, angSpeed := angSpeed(x = lon, y = lat, t = time, geo = TRUE)]
traj01
```

## Angular accelerations between successive steps

```{r}
traj01[, angAcc := angAcc(x = lon, y = lat, t = time, geo = TRUE)]
traj01
```

[Back to top]("#")

---

# Working with multiple trajectories

In order to simplify the manipulation of multiple trajectories, we will use the 
`dplyr` package from now on. 

```{r, message=FALSE}
library(dplyr)
```

## Load files in memory and compute all basic trajectory parameters

```{r}
filePaths <- dir(system.file("extdata", package = "swaRm"), full.names = TRUE)

trajs <- lapply(filePaths, function(path) {
  dat <- read.csv(path) 
  makeTraj(x = dat$lon, y = dat$lat, date = dat$date, time = dat$time,
           id = gsub(".*/|.csv.*", "\\1", path), geo = TRUE) %>%
    completeTraj(geo = TRUE) %>%
    fixTraj(geo = TRUE) %>%
    mutate(linDist = linDist(lon, lat, geo = TRUE)) %>%
    mutate(cumDist = cumsum(linDist)) %>%
    mutate(linSpeed = linSpeed(lon, lat, time, geo = TRUE),
           linAcc = linAcc(lon, lat, time, geo = TRUE),
           heading = heading(lon, lat, geo = TRUE),
           angSpeed = angSpeed(lon, lat, time, geo = TRUE),
           angAcc = angAcc(lon, lat, time, geo = TRUE))
}) %>% data.table::rbindlist()

trajs
```

## Representing the trajectories 

We will use `ggplot2` and `ggmap` to represent the different trajectories in 
their geographic context. 

```{r, cache=TRUE, message=FALSE}
library(ggplot2)
library(ggmap)

myLocation <- c(lon = mean(range(trajs$lon)), lat = mean(range(trajs$lat)))
myMap <- get_map(location = myLocation, source = "google", 
                 maptype = "satellite", zoom = 17, scale = 2)

ggmap(myMap) + 
  geom_path(data = trajs, aes(x = lon, y = lat, color = id), size = 0.8) +
  xlab("Longitude") + ylab("Latitude") +
  xlim(min(trajs$lon), max(trajs$lon)) + 
  ylim(min(trajs$lat), max(trajs$lat)) + 
  guides(color = FALSE)
```

## Calculating the centroid of the trajectories

```{r, warning=FALSE}
trajSummary <- group_by(trajs, time) %>%
  do(centroid(lon, lat, geo = TRUE))

ggmap(myMap) + 
  geom_path(data = trajs, aes(x = lon, y = lat, group = id), size = 0.8, color = "white") +
  geom_path(data = trajSummary, aes(x = lon, y = lat), size = 1, color = "red") +
  xlab("Longitude") + ylab("Latitude") +
  xlim(min(trajs$lon), max(trajs$lon)) + 
  ylim(min(trajs$lat), max(trajs$lat))
```

## Calculating the distances to the centroid of the trajectories

```{r}
trajs <- group_by(trajs, time) %>%
  mutate(distToCentroid = dist2centroid(lon, lat, geo = TRUE))

trajSummary <- group_by(trajs, time) %>% 
  summarize(distToCentroid = mean(distToCentroid)) %>%
  merge(trajSummary, by = "time")
```

## Determining the identity of and distance to the nearest neighbor

```{r}
trajs <- group_by(trajs, time) %>%
  mutate(nearNeighbor = nn(lon, lat, id, geo = TRUE),
         nearNeighborDist = nnd(lon, lat, geo = TRUE))
```

## Determining which individuals are on the convex hull of the group

```{r}
trajs <- group_by(trajs, time) %>%
  mutate(isChull = isChull(lon, lat))


# Plot the convex hull for a given timestamp
tmp <- filter(trajs, time == trajs$time[1]) 
tmp2 <- filter(tmp, isChull == TRUE) 
  
myLocation <- c(lon = mean(range(tmp$lon)), lat = mean(range(tmp$lat)))
myMap <- get_map(location = myLocation, source = "google", 
                 maptype = "satellite", zoom = 19, scale = 2)

ggmap(myMap) + 
  geom_point(data = tmp, aes(lon, lat), color = "red") + 
  xlab("Longitude") + ylab("Latitude") +
  xlim(min(tmp$lon), max(tmp$lon)) + 
  ylim(min(tmp$lat), max(tmp$lat)) + 
  guides(color = FALSE)
```

[Back to top]("#")

---







