---
title: "`swaRm` - a package for processing collective movement data"
author: "Simon Garnier - <a href='https://twitter.com/sjmgarnier'>@sjmgarnier</a>"
date: "`r Sys.Date()`"
output:
  html_document:
    css: custom.css
    fig_caption: true
    fig_height: 6
    fig_width: 8
    number_sections: true
    theme: cosmo
    highlight: pygments
    toc: true
    code_folding: show
  pdf_document:
    toc: true
vignette: |
  %\VignetteIndexEntry{Vignette Title} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

---

# Introduction

The goal of this vignette is to demonstrate the functionalities of the `swaRm` 
package for processing collective movement data. 

[Back to top]("#")

---

# Installation

`swaRm` can be installed from the [SwarmLab](http://www.theswarmlab.com)'s 
[`drat` repository](https://github.com/swarm-lab/drat) by running the following
lines in your R terminal: 

```{r, eval=FALSE, message=FALSE, warning=FALSE}
install.packages("drat")  # This line is required if drat is not present 
                          # in your R installation.

drat::addRepo("swarm-lab")
install.packages("swaRm")
```

If you want to receive package updates automatically, you can add the `drat` 
repository to your R installation permanently by adding the line 
`drat::addRepo("swarm-lab")` to your `.Rprofile` file.

[Back to top]("#")

---

# Preparing a trajectory data table

The `swaRm` package ships with a number of example data files that we will use 
for demonstrating its functioning. 

First, we will load in memory the first of these example data files. 

```{r, message=FALSE, warning=FALSE}
filePath <- system.file("extdata/01.csv", package = "swaRm")
dat <- read.csv(filePath)
head(dat)
```

This file contains GPS data, with a date, a timestamp, and longitude/latitude 
coordinates. Note that the coordinates can also be X/Y cartesian (or projected) 
coordinates (e.g. as returned by video tracking software, or if you projected 
your GPS data on a given reference grid).

The first step of the analysis process is to transform this imported data into 
a standardized trajectory data table that will be usable by all the functions in 
the `swaRm` package. 

```{r, message=FALSE, warning=FALSE, results='hide'}
library(swaRm)
traj <- makeTraj(x = dat$lon, y = dat$lat, id = "01", 
                 date = dat$date, time = dat$time, 
                 date.format = "ymd", time.format = "hms",
                 geo = TRUE)  # Set 'geo' to FALSE if working with cartesian
                              # (i.e. non geographic) coordinates
traj
```

Here `makeTraj` returns a table containing `r nrow(traj)` observations (each row
is an observation) and 4 columns: `id`, `time`, `lon` and `lat` (or `x` and `y` 
if working with projected data). `id` is a unique identifier chosen by you to 
distinguish between different trajectories when merging multiple tables together
(we will see this later in this tutorial). `time` is a standardized timestamp
constructed from the dates and times we provided the `makeTraj` function. 

Trajectory tables are special [`tibble` objects](https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html).
As such, they retain all the properties of `tibble` objects and can be 
manipulated in all the same ways `tibble` objects can be. This includes the 
possibility to use functions from the [`dplyr`](https://cran.r-project.org/package=dplyr) 
package as we will see later on. 

[Back to top]("#")

---

# Dealing with errors and missing observations

Errors and missing observations are common problems with tracking data, for 
instance when working in areas with poor GPS satellite covering. `swaRm` 
provides a set of convenience functions to automatically detect and correct some
of the most common types of errors. Detection and correction are handled by 
separate functions to allow users to develop their own correction functions for 
instance. 

These functions are meant to assist the user during the processing of his/her 
tracking data. However, as every automated method, they are not 100% reliable 
and a manual inspection of the data is recommended to ensure that all errors 
were correctly detected and corrected appropriately. It is also recommended that
you keep the original tracking data untouched in order to facilitate the 
comparison between the data sets before and after the correction is applied. 
Note as well that the order in which you apply these different corrections will 
have an impact on the final state of your data.

## Missing observations

Missing observations are probably the most common type of errors in tracking 
data. They can have a number of origins (e.g. GPS tag losing satellite 
connection, animal moving out of the camera field of view, read/write errors, 
etc.). The `findMissing` function is here to help you identify this missing data. 
Note that this function works better with data that was collected at regular 
time intervals. It will most likely return weird results if it is not the case. 

```{r, message=FALSE, warning=FALSE}
missing <- findMissing(traj)
missing
```

`findMissing` returns a data frame with two columns. The `time` column contains
timestamps at which the function believes observations are missing. The `type`
column indicates whether the observation is truly missing (i.e. was never 
recorded) or if the timestamp of the observation was set to NA during recording
or during the importation of the data (e.g. because the date or time info were
not formatted properly). You can then decide what to do about this missing data: 
ignore it, correct it with your own algorithm, or use the automated `fixMissing` 
function provided with the `swaRm` package. 

```{r, message=FALSE, warning=FALSE}
traj01 <- fixMissing(traj) 
traj01[1:40, ]
```

Here the function added two new observations at rows 3 and 4 (indicated by the 
`MISSING` error tag) and fixed to NA timestamps at rows 26 and 36 (indicated by 
the `NA` error tag).

Note that all error correction function will add the extra `error` column to the 
trajectory table in order to indicate which observations were corrected. This 
can be useful for instance to check that the automated correction did work 
properly.

## Duplicated timestamps

Duplicated timestamps are a less common type of errors. They can be caused by 
writing errors for instance. The function `findTimeDup` is here to help you 
identify these potential errors. 

```{r, message=FALSE, warning=FALSE}
time_dup <- findTimeDup(traj01) 
time_dup
```

The function `fixTimeDup` will then attempt to automatically correct the 
duplicated timestamp. If a timestamps is missing where the duplicated timestamp 
is, the function will replace the duplicated timestamp with the missing one. 
Otherwise, the duplicated timestamp will be replaced by NA. 

```{r, message=FALSE, warning=FALSE}
traj01 <- fixTimeDup(traj01)
traj01[1:20, ]
```

Here the duplicated timestamp corresponds to an already existing observation 
(hence the `<NA>` created by the `fixTimeDup` function in the `time` column to 
replace the duplicated timestamp). We will simply remove it here using the 
`filter` function in the `dplyr` package. 

```{r, message=FALSE, warning=FALSE}
library(dplyr)
traj01 <- filter(traj01, error != "TIMEDUP")
traj01[1:20, ]
```

## Inconsistent locations

Inconsistent locations are another frequent type of errors in tracking data. 
They correspond to reported locations that are significantly off compared to the 
normal variability of the tracking systems. They can have multiple origins (e.g.
GPS temporarily losing connection to one or more satellites, video recording 
showing increased noise, etc.) and they can create serious problems during the 
analysis of the data. For instance, here is a graph of the trajectory that we 
have created earlier. 

```{r, message=FALSE, warning=FALSE}
library(ggplot2)

ggplot(traj01, aes(x = lon, y = lat)) +
  geom_path() + 
  xlab("Longitude") + ylab("Latitude") +
  coord_map()
```

This does not look very good because one data point is way out of the actual
range of the data. We will use the `findLocErr` function to identify this point
and filter it out of the data when plotting. 

```{r, message=FALSE, warning=FALSE}
loc_err <- findLocErr(traj01) 
loc_err

ggplot(traj01[-loc_err, ], aes(x = lon, y = lat)) +
  geom_path() + 
  xlab("Longitude") + ylab("Latitude") +
  coord_map()
```

This is much better! 

Instead of filtering out the erroneous point, we can correct it using the 
`fixLocErr` function that will estimate its location using linear (default) or 
spline interpolation. 

```{r, message=FALSE, warning=FALSE}
traj01 <- fixLocErr(traj01) 

ggplot(traj01, aes(x = lon, y = lat)) +
  geom_path() + 
  xlab("Longitude") + ylab("Latitude") +
  coord_map()
```

Interpolation should be used carefully as it will introduce artificial 
auto-correlation between successive data points. If this is probably fine to 
correct spurious errors, it might create more serious problem when several 
successive locations require correction. In this case, it might be better to 
simply filter out the errors or ignore the corresponding segment in the
analyis. 

## NA locations

In some occasions, the data can contain timestamps that have no location data
associated with. This can occur for instance because a GPS unit could not 
acquire statellite signals. The `fixMissing` function seen earlier will also 
introduce such "NA" locations when adding missing timestamps to the trajectory
table. You can discover these missing locations in your data by using the 
`findLocNA` function.

```{r, message=FALSE, warning=FALSE}
loc_NA <- findLocNA(traj01)
loc_NA
```

You can then automatically replace these missing data points using the `fixLocNA`
function. This function will estimate the location of the missing points using 
linear (default) or spline interpolation. 

```{r, message=FALSE, warning=FALSE}
traj01 <- fixLocNA(traj01)
traj01[1:20, ]
```

As said before, interpolation should be used carefully as it will introduce 
artificial auto-correlation between successive data points. If this is probably 
fine to correct spurious errors, it might create more serious problem when 
several successive locations require correction. In this case, it might be 
better to simply filter out the errors or ignore the corresponding segment in 
the analyis. 

[Back to top]("#")

---

# Representing a geographic trajectory table using ggplot2 and ggmap

We have already seen earlier how to represent a trajectory using `ggplot2`. For 
geographic data, we can go one step further and use the 
[`ggmap`](https://cran.r-project.org/package=ggmap) package to download a 
satellite map of the study area and overlay the trajectory directly over it. 

```{r, message=FALSE, warning=FALSE}
library(ggmap)

myLocation <- c(lon = mean(range(traj01$lon)), lat = mean(range(traj01$lat)))
myMap <- get_map(location = myLocation, source = "google", 
                 maptype = "satellite", zoom = 17, scale = 2)

ggmap(myMap) + 
  geom_path(data = traj01, aes(x = lon, y = lat), color = "#D7601C", size = 1) +
  xlab("Longitude") + ylab("Latitude") +
  xlim(min(traj01$lon), max(traj01$lon)) + 
  ylim(min(traj01$lat), max(traj01$lat)) + 
  guides(color = FALSE)
```

[Back to top]("#")

---

# Computing the basic parameters of a trajectory

`swaRm` contains a number of functions to help you characterize the trajectories
that you are working with. We will go over these different functions in the rest
of this section. Most of them should be self-explanatory and I will therefore 
only provide example code. Note that I will be using the `mutate` function of 
the `dplyr` package in order to quickly append the calculated statistics to the
trajectory table.

## Linear and cumulative distances between successive steps

```{r, message=FALSE, warning=FALSE}
traj01 <- mutate(traj01, 
                 linDist = linDist(x = lon, y = lat, geo = is.geo(traj01)),
                 cumDist = cumsum(linDist))
traj01
```

Note that `geo` parameter of the `linDist` function is set to `TRUE` by the 
`is.geo` function in order to tell the `linDist` function that it will be 
handling geographic data. In this case, it uses the `distGeo` function from the
[`geosphere`](https://cran.r-project.org/package=geosphere) package to return 
the distances in meters. If `geo` had been set to `FALSE` instead, the returned 
distances would have been in the units of the data. The same principle will 
apply for the functions calculating linear speeds and accelerations below.

## Linear speeds between successive steps

```{r, message=FALSE, warning=FALSE}
traj01 <- mutate(traj01, 
                 linSpeed = linSpeed(x = lon, y = lat, t = time, geo = is.geo(traj01)))
traj01
```

## Linear accelerations between successive steps

```{r, message=FALSE, warning=FALSE}
traj01 <- mutate(traj01,
                 linAcc = linAcc(x = lon, y = lat, t = time, geo = is.geo(traj01)))
traj01
```

## Headings between successive steps

```{r, message=FALSE, warning=FALSE}
traj01 <- mutate(traj01,
                 heading = heading(x = lon, y = lat, geo = is.geo(traj01)))
traj01
```

As before, note that `geo` parameter of the `heading` function is set to `TRUE` 
by the `is.geo` function in order to tell the `heading` function that it will be 
handling geographic data. In this case, it uses the `bearing` function from the 
`geosphere` package to return the headings in radians. If `geo` had been set to 
`FALSE` instead, the heading would have been calculated using the `atan2` 
function and returned in radians as well. The same principle will apply for the 
functions calculating angular speeds and accelerations below.

## Angular speeds between successive steps

```{r, message=FALSE, warning=FALSE}
traj01 <- mutate(traj01,
                 angSpeed = angSpeed(x = lon, y = lat, t = time, geo = is.geo(traj01)))
traj01
```

## Angular accelerations between successive steps

```{r, message=FALSE, warning=FALSE, results='hide'}
traj01 <- mutate(traj01,
                 angAcc = angAcc(x = lon, y = lat, t = time, geo = is.geo(traj01)))
traj01
```

[Back to top]("#")

---

# Working with multiple trajectories

While you can use `swaRm` to process and analyze single trajectories, its main 
goal is to deal with multiple trajectories at once in order to calculate 
statistics about the collective behavior of animal moving in groups. In this 
section we will explore the different functions that this package offers to do 
exactly this. 

## Load multiple files in memory and compute all basic trajectory parameters

The first step will be to load all the sample trajectories that are provided 
with the `swaRm` package. There are 16 of them that correspond to the trajectories
of 16 animals collected using GPS collars. We will also compute all the basic 
statistics of these trajectories using the functions that we have explored in 
the previous section. We will use the `lapply` function to create and process 
all the trajectory tables separately and then the `rbindtt` function to combine 
them all together. This function is a wrapper around the `rbindlist` function of 
the `data.table` package to make it work properly with trajectory tables.

```{r, message=FALSE, warning=FALSE}
filePaths <- dir(system.file("extdata", package = "swaRm"), full.names = TRUE)

trajs <- lapply(filePaths, function(path) {
  dat <- read.csv(path) 

  makeTraj(x = dat$lon, y = dat$lat, id = gsub(".*/|.csv.*", "\\1", path),
           date = dat$date, time = dat$time, date.format = "ymd", time.format = "hms",
           geo = TRUE) %>%
    fixMissing() %>% 
    fixTimeDup() %>% filter(!is.na(time)) %>%
    fixLocErr() %>% 
    fixLocNA() %>%
    mutate(linDist = linDist(lon, lat, geo = TRUE)) %>%
    mutate(cumDist = cumsum(linDist)) %>%
    mutate(linSpeed = linSpeed(lon, lat, time, geo = TRUE),
           linAcc = linAcc(lon, lat, time, geo = TRUE),
           heading = heading(lon, lat, geo = TRUE),
           angSpeed = angSpeed(lon, lat, time, geo = TRUE),
           angAcc = angAcc(lon, lat, time, geo = TRUE))
}) %>% bind_rows()

trajs
```

Note the use of the pipe symbol (`%>%`) from the `dplyr` package to create a 
data processing pipeline. Read more about `dplyr` and `%>%` 
[here](http://stat545.com/block009_dplyr-intro.html).  

## Representing the trajectories using ggplot2 and ggmap

```{r, message=FALSE, warning=FALSE}
myLocation <- c(lon = mean(range(trajs$lon)), lat = mean(range(trajs$lat)))
myMap <- get_map(location = myLocation, source = "google", 
                 maptype = "satellite", zoom = 17, scale = 2)

ggmap(myMap) + 
  geom_path(data = trajs, aes(x = lon, y = lat, color = id), size = 0.8) +
  xlab("Longitude") + ylab("Latitude") +
  xlim(min(trajs$lon), max(trajs$lon)) + 
  ylim(min(trajs$lat), max(trajs$lat)) + 
  guides(color = FALSE)
```

## Summarizing group properties

Now that we have loaded and cleaned up the trajectories, we can start working
toward extracting the characteristics of the group behavior. `swaRm` provides a
number of functions to compute the most common statistics of group movement. 
More will be added in the future as the package keeps being developed. 

Most of the functions demonstrated below should be self-explanatory and I will 
therefore only provide example code. Note that I will be using functions of the 
`dplyr` package in order to facilitate the processing of the data over the 
entire trajectory table.

### Calculating the centroid of the trajectories

```{r, message=FALSE, warning=FALSE}
trajSummary <- group_by(trajs, time) %>%
  do(centroid(.$lon, .$lat, geo = TRUE)) %>%
  ungroup()

trajSummary

ggmap(myMap) + 
  geom_path(data = trajs, aes(x = lon, y = lat, group = id), size = 0.8, color = "white") +
  geom_path(data = trajSummary, aes(x = lon, y = lat), size = 1, color = "red") +
  xlab("Longitude") + ylab("Latitude") +
  xlim(min(trajs$lon), max(trajs$lon)) + 
  ylim(min(trajs$lat), max(trajs$lat))
```

### Calculating the distances to the centroid of the trajectories

```{r, message=FALSE, warning=FALSE}
trajs <- group_by(trajs, time) %>%
  mutate(distToCentroid = dist2centroid(lon, lat, geo = TRUE)) %>%
  ungroup()

trajs

trajSummary <- group_by(trajs, time) %>% 
  summarize(distToCentroid = mean(distToCentroid)) %>%
  merge(trajSummary, by = "time") 

trajSummary

ggplot(trajSummary, aes(x = time, y = distToCentroid)) +
  geom_line() + 
  xlab("Time") + ylab("Distance to the group's centroid (m)") + 
  ylim(0, NA)
```

### Determining the identity of and distance to the nearest neighbor

```{r, message=FALSE, warning=FALSE}
trajs <- group_by(trajs, time) %>%
  mutate(nearNeighbor = nn(lon, lat, id, geo = TRUE),
         nearNeighborDist = nnd(lon, lat, geo = TRUE))

trajs

trajSummary <- group_by(trajs, time) %>% 
  summarize(nearNeighborDist = mean(nearNeighborDist)) %>%
  merge(trajSummary, by = "time")

trajSummary

ggplot(trajSummary, aes(x = time, y = nearNeighborDist)) +
  geom_line() + 
  xlab("Time") + ylab("Mean distance to nearest neighbor (m)") + 
  ylim(0, NA)
```

### Determining which individuals are on the group's convex hull

```{r, message=FALSE, warning=FALSE}
trajs <- group_by(trajs, time) %>%
  mutate(isChull = isChull(lon, lat))

# Plot the convex hull for a randomly chose timestamp
randTime <- trajs$time[sample(1:nrow(trajs), 1)]
subTraj <- filter(trajs, time == randTime)
chullPol <- filter(subTraj, isChull > 0) %>%
  arrange(isChull)
  
myLocation <- c(lon = mean(range(subTraj$lon)), lat = mean(range(subTraj$lat)))
myMap <- get_map(location = myLocation, source = "google", 
                 maptype = "satellite", zoom = 19, scale = 2)

ggmap(myMap) + 
  geom_polygon(data = chullPol, aes(lon, lat), fill = "red", alpha = 0.25) +
  geom_point(data = subTraj, aes(lon, lat), size = 3, color = "white") + 
  geom_point(data = subTraj, aes(lon, lat, color = id), size = 2) + 
  xlab("Longitude") + ylab("Latitude") +
  xlim(min(subTraj$lon), max(subTraj$lon)) + 
  ylim(min(subTraj$lat), max(subTraj$lat)) + 
  guides(color = FALSE)
```

### Measuring the perimeter and surface area of the group's convex hull

```{r, message=FALSE, warning=FALSE}
trajSummary <- group_by(trajs, time) %>% 
  summarize(chullPerim = chullPerimeter(lon, lat, geo = TRUE),
            chullArea = chullArea(lon, lat, geo = TRUE)) %>%
  merge(trajSummary, by = "time")

ggplot(trajSummary, aes(x = time, y = chullPerim)) +
  geom_line() + 
  xlab("Time") + ylab("Perimeter of the group's convex hull (m)") + 
  ylim(0, NA)

ggplot(trajSummary, aes(x = time, y = chullArea)) +
  geom_line() + 
  xlab("Time") + ylab(bquote("Surface area of the group's convex hull (" * m ^ 2 * ")")) + 
  ylim(0, NA)
```

### Estimating the group's shape

The group's general shape is estimated by fitting a confidence ellipse on either 
the coordinates of all the group members, or only those belonging to the convex
hull of the group (we will use the first method here). 

From this ellipse, we can extract two shape parameters:

1. Sphericity: this is calculated as the ratio between the short and long axis 
of the ellipse. A ratio of 1 indicates a perfectly circular ellipse. Lower ratios
indicate that the ellipse is elongated in a particular direction. 

```{r, message=FALSE, warning=FALSE}
trajSummary <- group_by(trajs, time) %>% 
  summarize(sphericity = sphericity(lon, lat)) %>%
  merge(trajSummary, by = "time")

ggplot(trajSummary, aes(x = time, y = sphericity)) +
  geom_line() + 
  xlab("Time") + ylab("Group's sphericity") + 
  ylim(0, 1)
```

2. Stretch direction: this is calculated as the angle (in radians) between the 
major axis of the ellipse and the x-axis. It should always be in [-$\pi/2$,$\pi/2$].

```{r, message=FALSE, warning=FALSE}
trajSummary <- group_by(trajs, time) %>% 
  summarize(stretch = stretch(lon, lat)) %>%
  merge(trajSummary, by = "time")

ggplot(trajSummary, aes(x = time, y = stretch)) +
  geom_line() + 
  xlab("Time") + ylab("Group's sphericity") + 
  ylim(-pi / 2, pi / 2)
```

[Back to top]("#")

---

